---
title: "Week 1: Intro to demographic concepts"
author: "Monica Alexander"
date: "1/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Read in the data 
All these data come from the [UN World Population Prospects 2017](https://population.un.org/wpp/).

Packages:
```{r}
library(tidyverse)
library(here)
```


Population data:
```{r}
d_male <- read_csv(here("data", "WPP2017_POP_F15_2_ANNUAL_POPULATION_BY_AGE_MALE.csv"), skip = 12)
d_male$sex <- "M"
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)
d_female$sex <- "F"

d <- rbind(d_male, d_female)
rm(d_male, d_female)

d <- d %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`)

head(d)
```

Mortality data:

```{r}
d_male <- read_csv(here("data", "WPP2017_MORT_F04_2_DEATHS_BY_AGE_MALE.csv"), skip = 12)
d_male$sex <- "M"
d_female <- read_csv(here("data", "WPP2017_MORT_F04_3_DEATHS_BY_AGE_FEMALE.csv"), skip = 12)
d_female$sex <- "F"

dm <- rbind(d_male, d_female)
rm(d_male, d_female)

dm <- dm %>% 
  rename(region = `Region, subregion, country or area *`,
         year = Period) %>% 
  mutate(year = as.numeric(substr(year, 1, 4))) %>% 
  select(-Index, -Variant, -Notes, -`Country code`)

head(dm)

```

# Crude Rates

Calculate the crude death rates for Kenya and Canada in 2010

```{r}
# get total populations across all age and sex
total_pops <- d %>% 
  filter(region=="Kenya"|region=="Canada", year==2010) %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(region) %>% 
  summarise(total_pop = sum(pop, na.rm = T))

# get total deaths 

total_deaths <- dm %>% 
  filter(region=="Kenya"|region=="Canada", year==2010) %>% 
  gather(age, deaths, -region, -year, -sex) %>% 
  mutate(deaths = as.numeric(deaths)) %>% 
  group_by(region) %>% 
  summarise(total_deaths = sum(deaths, na.rm = T))

# join these

total_pops %>% 
  left_join(total_deaths) %>% 
  mutate(CDR = total_deaths/total_pop)

```


# Population Pyramids

Population age structures of Kenya and Canada in 2015:

```{r}
d_long <- d %>% 
  gather(age_group, population, -region, -year, -sex) %>% 
  mutate(population = as.numeric(population),
         age = ifelse(age_group=="80+", 80.5, ifelse(age_group=="100+", 100, as.numeric(age_group))))


d_long %>% 
  filter(region=="Kenya"|region=="Canada", year == 2015, age_group!="80+") %>% 
  mutate(population=ifelse(sex=="M", -population, population)) %>% 
ggplot(aes(x = age, y = population, fill = sex)) + 
  facet_wrap(~region)+
  geom_bar(stat="identity")+
  ggtitle("Population in each age group")+
  ylab("Population")+
  coord_flip() + 
  scale_y_continuous(breaks = seq(-4000, 4000, 1000), 
                     labels = c(seq(4000, 0, -1000), seq(1000, 4000, 1000))) + 
  scale_fill_brewer(palette = "Set1") 
ggsave(here("plots", "CAN_KEN_pyramid.pdf"))

```

Change in age structures over time:

```{r}
d_long %>% 
  filter(region=="Kenya"|region=="Canada", year%in% c(1960, 1990, 2015), age_group!="80+") %>% 
  mutate(population=ifelse(sex=="M", -population, population)) %>% 
ggplot(aes(x = age, y = population, fill = sex)) + 
  facet_grid(year~region)+
  geom_bar(stat="identity")+
  ggtitle("Population in each age group")+
  ylab("Population")+
  coord_flip() + 
  scale_y_continuous(breaks = seq(-4000, 4000, 1000), 
                      labels = c(seq(4000, 0, -1000), seq(1000, 4000, 1000))) + 
  scale_fill_brewer(palette = "Set1") 
ggsave(here("plots", "CAN_KEN_pyramid_time.pdf"))
```


# Age-specific rates

Create and plot age-specific mortality rates (across both sexes)

```{r}

pops <- d %>% 
  filter(region=="Kenya"|region=="Canada") %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(region, year, age) %>% 
  summarise(pop = sum(pop)) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<95)

deaths <- dm %>% 
  filter(region=="Kenya"|region=="Canada") %>% 
  gather(age, deaths, -region, -year, -sex) %>% 
  mutate(deaths = as.numeric(deaths)) %>% 
  group_by(region, year, age) %>% 
  summarise(deaths = sum(deaths)) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<95)

# join these two tibbles and calculate rates

asmr <- pops %>% 
  left_join(deaths) %>% 
  mutate(mx = deaths/pop)

head(asmr)

```
Plot these for selected years
```{r}

asmr %>% 
  filter(year %in% c(1980, 1990, 2000,2010)) %>% 
  ggplot(aes(age, mx, color = region)) +
  geom_line() + 
  facet_grid(~year) + 
  scale_y_log10()
```

# Age-standardized rates

What would Kenya mortality look like in 2010 with Canada's age structure?

```{r}
kenya_2010 <- asmr %>% 
  ungroup() %>% 
  filter(year==2010, region=="Kenya")  %>% 
  rename(kpop = pop, kdeath = deaths, kmx = mx) %>% 
  select(-region)

canada_2010 <- asmr %>% 
  ungroup() %>% 
  filter(year==2010, region=="Canada")  %>% 
  rename(cpop = pop, cdeath = deaths, cmx = mx) %>% 
  select(-region, -age, -year)

kc_2010 <- bind_cols(kenya_2010, canada_2010)

# now calculate age-standardized rates using canada's population

kc_2010 %>% 
  mutate(std_deaths_kenya = cpop*kmx,
         std_deaths_canada = cpop*cmx) %>% 
  summarise(std_rate_kenya = sum(std_deaths_kenya)/sum(cpop),
            std_rate_canada = sum(std_deaths_canada)/sum(cpop))

```


# Population Growth

Plot Kenya total population over time

```{r}
kenya_pop <- d %>% 
  filter(region=="Kenya") %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(year) %>% 
  summarise(pop = sum(pop, na.rm = T))

ggplot(kenya_pop, aes(year, pop)) + 
  geom_line() + 
  ggtitle("Kenya population over time")
```
What does the log look like?
```{r}

ggplot(kenya_pop, aes(year, log(pop))) + 
  geom_line() + 
  ggtitle("Kenya population (logged) over time")
ggsave(here("plots", "KEN_pop.pdf"))
```

Pretty straight! Let's calculate the growth rate from 1950 to 2015. This is just the slope of the logged graph. It's about 3% per year.  

```{r}
kenya_pop %>% 
  mutate(log_pop = log(pop)) %>% 
  summarise(growth_rate = (log_pop[year==2015] - log_pop[year==1950])/(2015-1950))
```

What about Canada? About half that. 

```{r}
canada_pop <- d %>% 
  filter(region=="Canada") %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(year) %>% 
  summarise(pop = sum(pop, na.rm = T))

canada_pop %>% 
  mutate(log_pop = log(pop)) %>% 
  summarise(growth_rate = (log_pop[year==2015] - log_pop[year==1950])/(2015-1950))
```

Some countries have stagnated:

```{r}
d %>% 
  filter(region=="Japan") %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(year) %>% 
  summarise(pop = sum(pop, na.rm = T)) %>% 
  mutate(log_pop = log(pop))  %>% 
  ggplot(aes(year, log_pop)) + geom_line()+
  ggtitle("Japan population (logged) over time")
```



# Extra: fertility charts in lecture notes

Read in data and tidy up

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)))

```

Plot age curves

```{r}
df %>% 
  filter(region=="Canada"|region=="Kenya", year==1960|year==2010) %>%
  gather(age, asfr, -region, -period, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  ggplot(aes(age, asfr/1000, color = region)) + 
  geom_point() + geom_line() + facet_grid(~year) +
  ggtitle("Age-specific fertility curves") + 
  ylab("ASFR (births/women)")
ggsave(here("plots", "CAN_KEN_fertility.pdf"))
```

