---
title: "Week 2 exercises with answers"
author: "Monica Alexander"
date: "1/18/2019"
output: html_document
---

# Read data  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
Packages:
```{r}
library(tidyverse)
library(here)
```

For this example we are using mortality in Canada by cause, 2012-2016. Available from [StatCan](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310039201). 

```{r}
d <- read_csv(here("data", "CAN_age_cod.csv"))
head(d)
```

# Mortality trends over time

What's happening to mortality rates over the five years? Plot the (log) age-specific mortality curves for all causes of death:

```{r}
 d %>% 
  filter(cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(year, age, Mx) %>% 
  ggplot(aes(age, log(Mx), color = factor(year))) + geom_line()
```


# Create a life table

Let's create a life table using the mortality rates from all causes of death in 2016. Get the data we need:

```{r}
dl <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, Mx)

head(dl)
```

You need to create columns:

- $n$
- $_na_x$
- $_nq_x$
- $_nd_x$
- $_nL_x$
- $T_x$
- $e_x$

What's the life expectancy at age 10?

```{r}
dl %>%
     mutate(n = case_when(
       age==0 ~ 1,
       age==1 ~ 4,
       TRUE ~ 5
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx,
       age==1 ~ 1.5,
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      qx = n * Mx / (1 + (n - ax)* Mx),
      px = 1 - qx,
      lx = lag(cumprod(px), default = 1),
      dx = lx - lead(lx, default = 0),
      Lx = n * lead(lx, default = 0) + (ax* dx),
      Tx = rev(cumsum(rev(Lx))),
      ex = Tx / lx
      ) 
```

# Calculate cause-deleted life expectancy

Now calculate life expectancy if all intentional injuries were deleted. What's the cause deleted life expectancy at age 10? What's the implied life lost due to suicide? Get the data we need:

```{r}
dls <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]"|cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, cause, Mx) %>% 
  mutate(cause = ifelse(cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]", "suicide", "total")) %>% 
  spread(cause, Mx) %>% 
  rename(Mx_i = suicide,
         Mx = total)

head(dls)
```

You need to create the same columns as above, but with the cause-deleted versions ($-i$). Do this by first creating the ratio $R_x^{-i} = \frac{M_x^{-i}}{M_x}$, use this to get $_nq_x^{-i}$, and the rest is the same. 

```{r}
dls %>%
     mutate(n = case_when(
       age==0 ~ 1,
       age==1 ~ 4,
       TRUE ~ 5
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx,
       age==1 ~ 1.5,
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      Rx = (Mx - Mx_i)/Mx,
      qx = n * Mx / (1 + (n - ax)* Mx),
      qxi = qx*Rx,
      px = 1 - qxi,
      lx = lag(cumprod(px), default = 1),
      dx = lx - lead(lx, default = 0),
      Lx = n * lead(lx, default = 0) + (ax* dx),
      Tx = rev(cumsum(rev(Lx))),
      ex = Tx / lx
      ) 
```


# Fit Gompertz model over time and plot parameters

Fit a Gompertz model to the total mortality rates for each year and plot the change over time. 

```{r}

gomp <- d %>% 
  filter(cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(year, age, Mx) %>% 
  mutate(log_mx = log(Mx)) %>% 
  filter(age > 39) %>% 
  group_by(year) %>% 
  summarise(alpha = exp((lm(log_mx~age))$coefficients[1]), beta = (lm(log_mx~age))$coefficients[2])

gomp %>% 
  ggplot(aes(alpha, beta, color = year)) + geom_point() +
  ggtitle("Gompertz parameters for Canada, 2012-2016")
```

How do these fit?

```{r}
d %>% 
  filter(cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(year, age, Mx) %>% 
  mutate(log_mx = log(Mx)) %>% 
  filter(age > 39) %>% 
  left_join(gomp) %>% 
  mutate(fit = log(alpha) + beta*age) %>% 
  ggplot(aes(age, log_mx)) + geom_point() + facet_grid(~year)  + geom_line(aes(age, fit))
```

