---
title: "Week 2: plots from lecture"
author: "Monica Alexander"
date: "1/18/2019"
output: html_document
---

The code for generating the plots that were shown in the week 2 lecture notes. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Read in the data 

Packages:
```{r}
library(tidyverse)
library(here)
```

Data: I found out about the Canadia HMD the other day! http://www.bdlc.umontreal.ca/CHMD/. Let's use the Ontario data.

```{r}
dm <- read_table(here("data", "ON_bltper_1x1.txt"), skip = 2)
head(dm)
```

## Characteristics life table function shapes

### Survivorship

```{r}
dm %>% 
  filter(Year==1921) %>% 
  select(Age, lx) %>% 
  ggplot(aes(as.numeric(Age), lx/100000)) + geom_line(lwd = 1.1) + 
  ylab("lx") + xlab("age") + 
  ggtitle("Survivorship curve for Ontario, 1921") 
ggsave(here("plots", "ON_lx_1921.pdf"))
```

```{r}
dm %>% 
  filter(Year==1921|Year==2011) %>% 
  mutate(Year = factor(Year)) %>% 
  ggplot(aes(as.numeric(Age), lx/100000, color = Year)) + geom_line(lwd = 1.1) + 
  ylab("lx") + xlab("age") + 
  ggtitle("Survivorship curve for Ontario, 1921 and 2011") 
ggsave(here("plots", "ON_lx_1921_2011.pdf"))
```
### Probability of dying

```{r}
dm %>% 
  filter(Year==1921|Year==2011) %>% 
  mutate(Year = factor(Year)) %>% 
  ggplot(aes(as.numeric(Age), qx, color = Year)) + geom_line(lwd = 1.1) + 
  ylab("qx") + xlab("age") + 
  ggtitle("Probability of dying, Ontario, 1921 and 2011") + 
  scale_y_log10()
ggsave(here("plots", "ON_qx_1921_2011.pdf"))
```
### Life expectancy
```{r}
dm %>% 
  filter(Year==1921|Year==2011) %>% 
  mutate(Year = factor(Year)) %>% 
  ggplot(aes(as.numeric(Age), ex, color = Year)) + geom_line(lwd = 1.1) + 
  ylab("qx") + xlab("age") + 
  ggtitle("Life expectancy, Ontario, 1921 and 2011") 
ggsave(here("plots", "ON_ex_1921_2011.pdf"))
```

## Gompertz models

## Plot the data

```{r}
dm %>% 
  filter(Year==1921|Year==2011) %>% 
  mutate(Year = factor(Year)) %>% 
  ggplot(aes(as.numeric(Age), mx, color = Year)) + geom_line(lwd = 1.1) + 
  ylab("qx") + xlab("age") + 
  ggtitle("Probability of dying, Ontario, 1921 and 2011") + 
  scale_y_log10()
ggsave(here("plots", "ON_mx_1921_2011.pdf"))
```

### Fit Gompertz

```{r}
dm %>% 
  mutate(age = as.numeric(Age)) %>% 
  filter(age>49) %>% 
  select(Year, age, mx) %>% 
  mutate(log_mx = log(mx)) %>% 
  group_by(Year) %>% 
  summarise(alpha = exp((lm(log_mx~age))$coefficients[1]), beta = (lm(log_mx~age))$coefficients[2]) %>% 
  ggplot(aes(alpha, beta, color = Year)) + geom_point() +
  ggtitle("Gompertz parameters for Ontario, 1921 - 2011")
ggsave(here("plots", "ON_gompertz.pdf"))
```

## Evidence of a mortality plateu?

Read in data for Sweden. Note that Sweden has exceptionally good-quality data going back a long time so is often used in examples. 

```{r}
ds <- read_table(here("data", "SWE_Mx_1x1.txt"), skip = 2)
head(ds)
```

```{r}
ds %>% 
  filter(Year==2015|Year==1800) %>% 
  mutate(age = as.numeric(Age), mx = as.numeric(Total), year = as.factor(Year)) %>% 
  filter(age>59) %>% 
  ggplot(aes(age, log(mx), color = year)) + geom_point() + 
  ggtitle("Log mortality rates, Sweden")
ggsave(here("plots", "SWE_log_mx.pdf"))
```

## Lee-carter model

Let's get the Lee-carter model parameters for Ontario. First, get the matrix of age-specific rates:

```{r}
m_tx <- dm %>% 
  select(Year, Age, mx) %>% 
  mutate(age = as.numeric(Age)) %>% 
  filter(age < 101) %>% 
  select(-Age) %>% 
  spread(age, mx) %>% 
  select(-Year) %>% 
  as.matrix()

ages <- 0:100
years <- unique(dm$Year)
```

log and demean those rates:

```{r}

logm_tx <- log(m_tx)
ax <- apply(logm_tx, 2, mean)
```

Do the SVD

```{r}
swept_logm_tx <- sweep(logm_tx, 2, ax)

svd_mx <- svd(swept_logm_tx)

bx <- svd_mx$v[, 1]/sum(svd_mx$v[, 1])
kt <- svd_mx$d[1] * svd_mx$u[, 1] * sum(svd_mx$v[, 1])

```

Plots!

```{r}
# plot ax

lc_age_df <- tibble(age = ages, ax = ax, bx = bx)
lc_time_df <- tibble(year = years, kt = kt)

ggplot(lc_age_df, aes(age, ax)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("ax values for Ontario")
ggsave(here("plots", "ON_ax.pdf"))

ggplot(lc_age_df, aes(age, bx)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("bx values for Ontario")
ggsave(here("plots", "ON_bx.pdf"))

ggplot(lc_time_df, aes(year, kt)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("kt values for Ontario")
ggsave(here("plots", "ON_kt.pdf"))
```

let's look at the fit for a couple of years

```{r}
data_and_res <- dm %>% 
  select(Year, Age, mx) %>% 
  mutate(age = as.numeric(Age), year = Year) %>% 
  filter(age < 101) %>% 
  select(-Age, -Year) %>% 
  mutate(log_mx = log(mx)) %>% 
  left_join(lc_age_df) %>% 
  left_join(lc_time_df) %>% 
  mutate(lc_fit = ax + bx*kt)

data_and_res %>% 
  filter(year %in% c(1921, 1951, 1981, 2011)) %>% 
  ggplot(aes(age, log_mx)) + geom_point() + facet_grid(~year) + 
  geom_line(aes(age, lc_fit), color = "red") + 
  ggtitle("Data and fits for Ontario")
ggsave(here("plots", "ON_lc.pdf"))

```

