---
title: "Week 2 exercises"
author: "Monica Alexander"
date: "1/18/2019"
output: html_document
---

A short lab to calculate life tables, a life table with suicide removed, and Gompertz parameters for recent mortality in Canada. 

# Read data  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
Packages:
```{r}
library(tidyverse)
library(here)
```

For this example we are using mortality in Canada by cause, 2012-2016. Available from [StatCan](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310039201). 

```{r}
d <- read_csv(here("data", "CAN_age_cod.csv"))
head(d)
```


# Create a life table

Let's create a life table using the mortality rates from all causes of death in 2016. Get the data we need:

```{r}
dl <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, Mx)

head(dl)
```

You need to create columns:

- $n$
- $_na_x$
- $_nq_x$
- $_nd_x$
- $_nL_x$
- $T_x$
- $e_x$

What's the life expectancy at age 10?

# Calculate cause-deleted life expectancy

Now calculate life expectancy if all intentional injuries were deleted. Get the data we need:

```{r}
dls <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]"|cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, cause, Mx) %>% 
  mutate(cause = ifelse(cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]", "suicide", "total")) %>% 
  spread(cause, Mx) %>% 
  rename(Mx_i = suicide,
         Mx = total)

head(dls)
```

You need to create the same columns as above, but with the cause-deleted versions ($-i$). Do this by first creating the ratio $R_x^{-i} = \frac{M_x^{-i}}{M_x}$, use this to get $_nq_x^{-i}$, and the rest is the same. 

What's the cause deleted life expectancy at age 10? What's the implied life lost due to suicide?


# Fit Gompertz model over time and plot parameters

Fit a Gompertz model to the total mortality rates for each year and plot the change over time. Check the fit by plotting data and fitted values. 

