---
title: 'Week 3: Fertility plots'
author: "Monica Alexander"
date: "1/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

Code to produce plots in lecture notes. 

## Data

```{r}
library(tidyverse)
library(here)
```
Read in data on CBR and CDR and get it into a workable format. 
```{r}

db <- read_csv(here("data", "WPP2017_FERT_F03_CRUDE_BIRTH_RATE.csv"), skip = 16)
dd <- read_csv(here("data", "WPP2017_MORT_F02_CRUDE_DEATH_RATE.csv"), skip = 16)

db <- db %>% 
  rename(region = `Region, subregion, country or area *`) %>% 
  select(region, `1950-1955`:`2010-2015`) %>% 
  gather(year, cbr, -region)

dd <- dd %>% 
  rename(region = `Region, subregion, country or area *`) %>% 
  select(region, `1950-1955`:`2010-2015`) %>% 
  gather(year, cdr, -region)

d <- db %>% 
  left_join(dd)

# change year to be numeric

d <- d %>% 
  rename(period = year) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)))

```

Read in the population data and get into a workable format

```{r}
d_male <- read_csv(here("data", "WPP2017_POP_F15_2_ANNUAL_POPULATION_BY_AGE_MALE.csv"), skip = 12)
d_male$sex <- "M"
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)
d_female$sex <- "F"

dp <- rbind(d_male, d_female)
rm(d_male, d_female)

dp <- dp %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`)

head(dp)
```

Calculate total population sizes

```{r}
d_tot <- dp %>% 
  gather(age, pop, -region, -year, -sex) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  group_by(region, year)  %>% 
  summarise(pop = sum(pop, na.rm = T))
```

Pick some countries and plot

```{r}
d %>% 
  left_join(d_tot) %>% 
  mutate(pop = pop / 2000) %>% 
  gather(measure, value, -region, -period, -year) %>% 
  filter(region %in% c("Nigeria", "Japan", "United Kingdom")) %>% 
  ggplot(aes(year, value, color = measure)) + 
  geom_line(lwd = 1.1) + 
  facet_grid(~region) +
  ylab("cdr, cbr, pop/2*10^8")+
  theme_gray(base_size = 14)+
  ggtitle("Birth rates, death rates and population, 1950-2015")
ggsave(here("plots", "JPN_NGA_UK_dem_transition.pdf"))
```

## Age-specific fertility rates

Read in WPP data

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)))
```

Plot age curves

```{r}
df %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Indonesia", "India", "Kenya")) %>%
  gather(age, asfr, -region, -period, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  ggplot(aes(age, asfr/1000, color = year, group = year)) + 
  geom_line() + facet_wrap(~region) +
  ggtitle("Age-specific fertility curves") + 
  ylab("ASFR (births/women)")
ggsave(here("plots", "grid_fertility.pdf"))
```

## Total fertility rates


```{r}
df %>% 
  gather(age, asfr, -region, -period, -year) %>% 
  group_by(region, year) %>% 
  summarise(tfr = 5*sum(asfr/1000)) %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Indonesia", "India", "Kenya")) %>% 
  ggplot(aes(year, tfr, color = region))  + 
  geom_line(lwd = 1.1) +
  scale_color_viridis_d()+
  geom_hline(yintercept = 2.1) + 
    theme_gray(base_size = 14) + 
  ggtitle("Total fertility rate for selected countries, 1950-2015")
ggsave(here("plots", "tfr_ts.pdf"))
```

Just Canada: data from the [Human Fertility Database](https://www.humanfertility.org/).

```{r}
dfc <- read_table(here("data", "CANtfrRRbo.txt"), skip = 2)

dfc %>% 
  select(Year, TFR) %>% 
  ggplot(aes(Year, TFR)) + 
  geom_line(lwd = 1.1) +
  ggtitle("TFR for Canada, 1944-2011") + 
  theme_grey(base_size = 14) 
ggsave(here("plots", "CAN_tfr.pdf"))
```


## Parity progression ratios

Canadian data from the [Human Fertility Database](https://www.humanfertility.org/).

```{r}
ppr <- read_table(here("data", "CANpprVHbo.txt"), skip = 2)
head(ppr)
```
Change into long format 

```{r}
ppr <- ppr %>% 
  gather(measure, ratio, -Cohort) %>% 
  mutate(parity =  as.numeric(substr(measure, 6, 6))) %>% 
  rename(cohort = Cohort) %>% 
  select(-measure)
```


Plot!

```{r}
ppr %>% 
  mutate(parity = factor(parity)) %>% 
  ggplot(aes(cohort, ratio, color = parity)) + 
  geom_line(lwd = 1.1) + 
  scale_color_viridis_d() + 
  ylab("PPR") + 
  theme_gray(base_size = 14) + 
  ggtitle("Parity Progression Ratios in Canada") 
ggsave(here("plots", "CAN_PPR.pdf"))
```


Can also loook at TFR by parity:
```{r}
dfc_long <- dfc %>% 
  select(Year:TFR5p) %>% 
  select(-TFR) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(tfr = value)

dfc_long %>% 
  mutate(parity = factor(parity)) %>% 
  ggplot(aes(Year, tfr, color = parity)) + 
  geom_line(lwd = 1.1) + 
  scale_color_viridis_d()
  
```



## Net reproduction ratios

Read in WPP life table data and clean up. 

```{r}

dl <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)))
  

```

Join on the fertility rates and calculate some NRRs

```{r}

ffab <- 0.4886

NRRs <- df %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  left_join(dl) %>% 
  mutate(prod = Fx/1000*Lx/10^5*ffab) %>% 
  group_by(region, year) %>% 
  summarise(NRR = sum(prod))

```

Plot!

```{r}
NRRs %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Indonesia", "India", "Kenya")) %>% 
  ggplot(aes(year, NRR, color = region))  + 
  geom_line(lwd = 1.1) +
  scale_color_viridis_d()+
  geom_hline(yintercept = 1) + 
  theme_gray(base_size = 14) + 
  ggtitle("NRRs for selected countries, 1950-2015")
ggsave(here("plots", "nrr_ts.pdf"))
```


## Coale and Trussel's M and m

Define our constant age schedules

```{r}
n_x <- tibble(age = seq(15, 45, by = 5), nx = c(0.36, 0.46, 0.431, 0.396, 0.321, 0.167, 0.024))
nu_x <- tibble(age = seq(15, 45, by = 5), nu = c(0,0,0.279, 0.667, 1.042, 1.414, 1.67)) 
```

Fit the model

```{r}

ct_res <- df %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  left_join(n_x) %>% 
  left_join(nu_x) %>% 
  mutate(log_ratio = log(Fx/1000/nx)) %>% 
  mutate(isna = is.infinite(log_ratio)) %>% 
  filter(!isna) %>% 
  group_by(region, year) %>% 
  summarise(big_m = exp((lm(log_ratio~nu))$coefficients[1]),  small_m = -1*(lm(log_ratio~nu))$coefficients[2]) 

```


Plot fits and data (does not do a good job).

```{r}

data_and_fits <- df %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  left_join(n_x) %>% 
  left_join(nu_x) %>% 
  left_join(ct_res) %>% 
  mutate(fit_original_scale = nx*exp(log(big_m) - small_m*nu)*1000,
         dependent_var = log(Fx/1000/nx),
         fit_log_scale = log(big_m) - small_m*nu)


data_and_fits %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Bangladesh", "India", "Kenya")) %>% 
  filter(year %in% c(1950, 2010)) %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(age, dependent_var, color = year, group = year)) + 
  facet_wrap(~region) + geom_point() + 
  geom_line(aes(age, fit_log_scale))


data_and_fits %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Bangladesh", "India", "Kenya")) %>% 
  filter(year %in% c(1950, 2010)) %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(age, Fx, color = year)) + 
  facet_wrap(~region) + geom_point() + 
  geom_line(aes(age, fit_original_scale))+
  theme_gray(base_size = 14) + 
  ggtitle("Data and Coale and Trussel fit, 1950 and 2010")
ggsave(here("plots", "ct_fit.pdf"))

```


Plot scatter over time

```{r}
ct_res %>% 
  filter(region %in% c("Canada", "Nigeria", "Japan", "Bangladesh", "India", "Kenya")) %>% 
  ggplot(aes(big_m, small_m, color = year)) + 
  facet_wrap(~region) + 
  geom_point() + 
  theme_grey(base_size = 14) + 
  ylab("level of fertility") + 
  xlab("level of fertility limitation") + 
  ggtitle("Coale and Trussel's M and m for selected countries \n1950-2010")
ggsave(here("plots", "ct_scatter.pdf"))
```

## Tempo adjusted TFR

Calculate tempo-adjusted TFRs by parity then sum. Note the Human Fertility Database has the data. 


```{r}
dmac <- read_table(here("data", "CANmabRRbo.txt"), skip = 2)

mab_tfr <- dmac %>% 
  select(Year:MAB5p) %>% 
  select(-MAB) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(mab = value) %>% 
  left_join(dfc_long) 

tfr_adj <- mab_tfr %>% 
  mutate(diff_mab = (mab - lag(mab))/2,
         tfr_dash = tfr/(1-diff_mab)) %>% 
  group_by(Year) %>% 
  summarise(unadjusted = sum(tfr),
            tempo_adjusted = sum(tfr_dash)) 

head(tfr_adj)

```

Plot!

```{r}
tfr_adj %>% 
  gather(tfr, value, -Year) %>% 
  ggplot(aes(Year, value, color = tfr)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("TFR and tempo-adjusted TFR \nCanada 1944-2011")
ggsave(here("plots", "CAN_tempo.pdf"))
```

What about Japan?


```{r}
dfj <- read_table(here("data", "JPNtfrRRbo.txt"), skip = 2)
dmaj <- read_table(here("data", "JPNmabRRbo.txt"), skip = 2)

dfj_long <- dfj %>% 
  select(Year:TFR5p) %>% 
  select(-TFR) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(tfr = value)

mab_tfr <- dmaj %>% 
  select(Year:MAB5p) %>% 
  select(-MAB) %>% 
  gather(measure, value, - Year) %>% 
  mutate(parity = as.numeric(substr(measure, nchar(measure) , nchar(measure)))) %>% 
  mutate(parity = ifelse(is.na(parity), 5, parity)) %>% 
  select(-measure) %>% 
  rename(mab = value) %>% 
  left_join(dfj_long) 

tfr_adj <- mab_tfr %>% 
  mutate(diff_mab = (mab - lag(mab))/2,
         tfr_dash = tfr/(1-diff_mab)) %>% 
  group_by(Year) %>% 
  summarise(unadjusted = sum(tfr),
            tempo_adjusted = sum(tfr_dash)) 

tfr_adj %>% 
  gather(tfr, value, -Year) %>% 
  ggplot(aes(Year, value, color = tfr)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("TFR and tempo-adjusted TFR \nJapan 1944-2011")

```

