---
title: 'Week 4: Stable populations and population projection'
author: "Monica Alexander"
date: "1/31/2019"
output: html_document
---

Code to produce plots in lecture notes for week 4. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r}
library(tidyverse)
library(here)
```

# Stable population

Simple illustration of stable population. Set up the constants:

```{r}
lx <- c(1, 0.6, 0.4, 0.2, 0.05, 0)
B <- 1000
r <- 0.1
```

Set up initial population until we have a full column

```{r}
c1 <- diag(lx*B)

for(i in 1:(nrow(c1)-1)){
  for(j in i:(ncol(c1)-1)){
    c1[i,j+1] <- c1[i,j]*exp(r)
  }
}

colnames(c1) <- paste0("P", 1:6)

dp <- as_tibble(cbind(lx, as_tibble(c1)))

```

Now do a few more years

```{r}
dp <- dp %>% 
  mutate(P7 = P6*exp(r),
         P8 = P7*exp(r), 
         P9 = P8*exp(r))
```

Plot these

```{r}
dp %>% 
  mutate(age = 0:5) %>% 
  gather(time, population, -age, -lx) %>% 
  mutate(year = as.numeric(substr(time, 2, 2))) %>% 
  group_by(year) %>% 
  mutate(prop = population/ sum(population)) %>% 
  filter(year > 4) %>% 
  ungroup() %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(x = age, y = population, fill = year)) + 
  geom_bar(stat="identity", position = "dodge") +  coord_flip() +
  ggtitle("Population by year and age group")
ggsave(here("plots", "stable_pop.pdf"))

dp %>% 
  mutate(age = 0:5) %>% 
  gather(time, population, -age, -lx) %>% 
  mutate(year = as.numeric(substr(time, 2, 2))) %>% 
  group_by(year) %>% 
  mutate(prop = population/ sum(population)) %>% 
  filter(year > 4) %>% 
  ungroup() %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(x = age, y = prop, fill = year)) + 
  geom_bar(stat="identity", position = "dodge") +  coord_flip() +
  ggtitle("Proportion by year and age group")
ggsave(here("plots", "stable_prop.pdf"))
```

# Trial values of $r$

Let's plot the Euler-Lotka equation on the log scale as a function of R. First, we need Lx and Fx values

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4))) %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)

dl <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), Lx = Lx/10^5)
```

Let's look at trial values of $r$ for Canada. The NRR is less than one, so what is $r$?

```{r}

df %>% 
  left_join(dl) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(prod = Fx*Lx*0.4886) %>% 
  summarise(sum(prod))

```

Let's plot $Y(r)$ at different $r$

```{r}

prod <- df %>% 
  left_join(dl) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(prod = Fx*Lx*0.4886) %>% 
  select(prod) %>% 
  pull()

ages <- seq(15, 45, by = 5)

get_Yr <- function(prod, r){
  exp_ra <- exp(-r*ages)
  return(sum(prod*exp_ra))
}

get_Yr(prod, r)

rs <- seq(-0.1, 0.1, by = 0.001)
Yrs <- sapply(1:length(rs), function(i) get_Yr(prod, rs[i]))

ggplot(data = tibble(r = rs, log_Yr = log(Yrs)), aes(r, log_Yr)) + 
  geom_line() + 
  geom_hline(yintercept = 0, lty = 2) + 
  geom_vline(xintercept = -0.009, color = "red") + 
  ggtitle("Log of Euler-Lotka for different values of r \nCanada 2010")
ggsave(here("plots", "euler_lotka_yr.pdf"))
```

# Leslie matrices

Create a function that makes a Leslie matrix based on $_nL_x$ and $_nF_x$ values

```{r}
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=17,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  L[1,ncol(L)] <- 0
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```

Make a Leslie matrix for Canada

```{r}
nLx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada", age<85) %>% 
  select(Lx) %>% pull()

## need to fix first age group

nLx <- c(sum(nLx[1:2]), nLx[3:length(nLx)])

nFx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx = ifelse(is.na(Fx), 0, Fx)) %>% 
  select(Fx) %>% pull()

nFx <- nFx[-1]

A <- leslie(nLx, nFx)
A
```

Note that we can calculate NRR from the matrix elements:

```{r}
A[1,]%*%cumprod(c(1,diag(A[-1,])))
```


## Projecting Canada's population

Get the population in 2010 from WPP data. 

```{r}
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)

Kt <- d_female %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  filter(region=="Canada", year==2010) %>% 
  gather(age, pop, -region, -year) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<85) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  select(pop) %>% 
  pull()
  
ggplot(tibble(age = seq(0, 80, by = 5), K = Kt), aes(age, K)) + 
  geom_bar(stat= "identity") + coord_flip() + ylab("Population") + ggtitle("Female population, Canada 2010")
ggsave(here("plots", "CAN_pop_2010.pdf"))
```

Do the projection up to 2050 in a loop:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 8
initial_pop <- Kt
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- A%*%K[,i-1] 
}

K
```

```{r}
Kdf <- as_tibble(K)
colnames(Kdf) <- seq(from = 2010, to = (2010+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 80, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
dk <- Kdf %>% 
  gather(year, population, -age) %>%
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>%
  mutate(proportion = population/sum(population))
dk
```

Plot total population size over time

```{r}
dk %>% 
  group_by(year) %>% 
  summarise(pop = sum(population)) %>% 
  ggplot(aes(year, pop)) + geom_line() + 
  ggtitle("Canada's female population, 2010-2050")
ggsave(here("plots", "CAN_pop_proj.pdf"))
```

Plot proportions over time

```{r}
dk %>% 
  filter(age %in% seq(0, 80, by = 10)) %>% 
  mutate(age = factor(age)) %>% 
  ggplot(aes(year, proportion, color = age)) + 
  geom_line() + 
  ggtitle("Proportion of population by age group \nCanada 2010-2050")
ggsave(here("plots", "CAN_prop_proj.pdf"))
```
Do the projection over a long time:

```{r}
n_projections <- 40
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- A%*%K[,i-1] 
}

Kdf <- as_tibble(K)
colnames(Kdf) <- seq(from = 2010, to = (2010+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 80, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
dk <- Kdf %>% 
  gather(year, population, -age) %>%
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>%
  mutate(proportion = population/sum(population))

dk %>% 
  filter(age %in% seq(0, 80, by = 10)) %>% 
  mutate(age = factor(age)) %>% 
  ggplot(aes(year, proportion, color = age)) + 
  geom_line() + 
  ggtitle("Proportion of population by age group \nCanada 2010-2250")
ggsave(here("plots", "CAN_prop_proj_long.pdf"))
```

## Eigendecomposition

```{r}

eigen_A <- eigen(A)
r_A <- log(Re(eigen_A$value[1]))/5 
v <- Re(eigen_A$vectors[,1])
k_A <- v/sum(v)

r_A
k_A

```


Reproductive value:

```{r}
eigen_A <- eigen(A[1:length(seq(0, 45, by = 5)), 1:length(seq(0, 45, by = 5))])
U <- eigen_A$vector
U <-U%*%diag(1/U[1,])
Uinv <- solve(U)
Vstar <- diag(1/Uinv[,1]) %*% Uinv # normalize

ggplot(tibble(age = seq(0, 45, by = 5), V = Re(Vstar[1,])), aes(age, V)) + 
  geom_line() + 
  ggtitle("Reproductive Values for Canada, 2010")
ggsave(here("plots", "CAN_rv.pdf"))
```

Mean age at childbearing:

```{r}
(Re((Vstar%*%U))*5)[1,1]
```

# Population momentum

Calculate the keyfitz scenario multiplier for Canada in 2010. Note, we need to read in ex values to get life expectancy:

```{r}
NRR <- sum(nFx*nLx*0.4886)
de <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

ex <- de %>% 
  rename(region = `Region, subregion, country or area *`,
         ex = `Expectation of life e(x)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, ex) %>% 
  mutate(year = as.numeric(substr(period, 1, 4))) %>% 
  filter(region == "Canada", age==0, year==2010) %>% 
  select(ex) %>% 
  pull()

b_pre <- 1/sum(exp(-r_A*seq(0, 80, by = 5))*nLx)

ex*b_pre/sqrt(NRR)

```

Let's actually make the Leslie matrix stationary and see what happens to population

```{r}
B <- rbind(A[1,]/NRR, A[-1,])

n_projections <- 40
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- B%*%K[,i-1] 
}

Kdf <- as_tibble(K)
colnames(Kdf) <- seq(from = 2010, to = (2010+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 80, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
dk <- Kdf %>% 
  gather(year, population, -age) %>%
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>%
  mutate(proportion = population/sum(population))

dk %>% 
  group_by(year) %>% 
  summarise(pop = sum(population)) %>% 
  ggplot(aes(year, pop)) + 
  geom_line() + 
  ggtitle("Population over time if NRR was 1 \nCanada 2010-2250")
ggsave(here("plots", "CAN_keyfitz_nonstable.pdf"))

dk %>% 
  group_by(year) %>% 
  summarise(pop = sum(population)) %>% 
  summarise(pop[row_number()==n()]/pop[row_number()==1])
```


# Extra stuff

Playing around with stationary and non-stationary momentum (not covered in lecture)

```{r}
U <- eigen(A[1:length(seq(0, 45, by = 5)), 1:length(seq(0, 45, by = 5))])$vectors
U <-U%*%diag(1/U[1,])
u <- as.vector(Re(U[,1]))
Uinv <- solve(U)
Vstar <- diag(1/Uinv[,1]) %*% Uinv 
v <- as.vector(Re(Vstar[1,]))

B <- rbind(A[1,]/NRR, A[-1,])
eigen_B <- eigen(B[1:length(seq(0, 45, by = 5)), 1:length(seq(0, 45, by = 5))])

U0 <- eigen_B$vector
U0 <-U0%*%diag(1/U0[1,])
u0 <- as.vector(Re(U0[,1]))
U0inv <- solve(U0)
Vstar0 <- diag(1/U0inv[,1]) %*% U0inv 
v0 <- as.vector(Re(Vstar0[1,]))

keyfitz <- sum(v0*u*5)/sum(v0*u0*5)*sum(u0)/sum(u)

kt <- Kt/sum(Kt)

stationfac <- sum(v0*kt*5)/sum(v0*u0*5)*sum(u0)/sum(kt)
stablefac <- sum(v*kt*5)/sum(v*u*5)*sum(u)/sum(kt)

stationfac
stablefac*keyfitz

```

