---
title: 'Week 5: Migration'
author: "Monica Alexander"
date: "2/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```


Projections with migration for Canada. 

# Load in WPP data

```{r}
library(tidyverse)
library(here)
```

Load in fertility rates and tidy up:

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4))) %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)

head(df)
```


Load in the life table data and tidy up:

```{r}
dl <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), Lx = Lx/10^5)

head(dl)
```


Load in the population data and tidy up: 

```{r}
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)

d_female <- d_female %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  gather(age, pop, -region, -year) %>% 
  mutate(age = as.numeric(age), pop = as.numeric(pop))

head(d_female)
```


# Canada migration data

For Canada we have some age-specific data from StatCan. This data gives the number of international migrants by age group. 

```{r}

dmig <- read_csv(here("data", "CAN_mig_age.csv"))

dmig <- dmig %>% mutate(mig = mig/1000)
head(dmig)
```

1. Plot migration by age as a proportion of total population. You will need to join the population (`d_female`) and migration (`dmig`) data sets.

# Building a Leslie Matrix

Let's create a Leslie Matrix for Canada using WPP data (like last week). Here's the `leslie` function:

```{r}
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=17,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  L[1,ncol(L)] <- 0
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```


Here's the matrix:

```{r}
nLx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada", age<85) %>% 
  select(Lx) %>% pull()

## need to fix first age group

nLx <- c(sum(nLx[1:2]), nLx[3:length(nLx)])

nFx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx = ifelse(is.na(Fx), 0, Fx)) %>% 
  select(Fx) %>% pull()

nFx <- nFx[-1]

A <- leslie(nLx, nFx)
A
```

# Do a base population projection (no migration yet)

Get the initial population vector for projection:

```{r}
Kt <- d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<85) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  select(pop) %>% 
  pull()
  
ggplot(tibble(age = seq(0, 80, by = 5), K = Kt), aes(age, K)) + 
  geom_bar(stat= "identity") + coord_flip() + ylab("Population") + ggtitle("Female population, Canada 2010")
```


Do a projection with this Leslie matrix for 40 periods:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- A%*%K[,i-1] 
}

```

# Add migration to the projection

2. Create a vector of migration rates by age, $I_t$. Note: you plotted this above. 

3. Repeat the projection for 40 periods, but this time with migration. 

Notes: 

- Save your projections as `K_mig` or something similar (don't overwrite `K` above)
- From the lecture 

\begin{eqnarray*}
K(t+5) &=&
A \times
\begin{bmatrix}
    _5K_0(t)+{_5K_0}(t){_5i_0} /2  \\
   _5K_5(t)+{_5K_5}(t){_5i_5} /2     \\
     _5K_10(t)+{_5K_10}(t){_5i_{10}} /2\\
    \dots \\
  _{\infty}K_{\omega}(t)+_{\infty}K_{\omega}(t){_5i_0} /2
\end{bmatrix}  + 
\begin{bmatrix}
    _5K_0(t){_5i_0} /2  \\
   _5K_5(t){_5i_5}  /2     \\
     _5K_{10}(t){_5i_{10}}/2 \\
    \dots \\
   _{\infty}K_{\omega}(t) _{\infty}i_{\omega}/2
\end{bmatrix}
\end{eqnarray*}

4. What difference does adding in migration make? Plot the age structure from the last projection time period for projections with and without migration. 

# Increasing fertility and migration

5. Calculate Canada's TFR. 

6. Construct a new set of age-fertility rates such that TFR = 2.5.

7. Construct a new Leslie matrix based on these higher fertility rates.

8. Do a projection for 40 steps based on this new Leslie matrix, and explore the result. Has the shape of the age pyramid changed?

9. Investigate what happens to the projected population for different migration scenarios (going back to the original fertility rates).
    + Does increasing migration proportional across all age groups change the shape of the age pyramid?
    + What is the implied increase in migration needed in ages < 50 such that the rate of increase of the population is the same as when you increased TFR to 2.5?
