---
title: 'Week 5: Migration'
author: "Monica Alexander"
date: "2/07/2019"
output: html_document
---

Code to produce plots in lecture notes for week 5. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r}
library(tidyverse)
library(here)
library(rstan)
```

# Load in data

Load in fertility rates and tidy up:

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4))) %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)

head(df)
```


Load in the life table data and tidy up:

```{r}
dl <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), Lx = Lx/10^5)

head(dl)
```

Load in the population data: 

```{r}
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)

d_female <- d_female %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  gather(age, pop, -region, -year)

head(d_female)
```

# Population pyramids for Canada

For Canada we have some age-specific data from StatCan:

```{r}

dmig <- read_csv(here("data", "CAN_mig_age.csv"))

dmig <- dmig %>% mutate(mig = mig/1000)
head(dmig)
```

Let's plot as a proportion of total population:

```{r}

d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age), pop = as.numeric(pop)) %>% 
  left_join(dmig) %>% 
  mutate(prop = mig/pop) %>% 
  ggplot(aes(age, prop)) + geom_line() + 
  ggtitle("Migration into Canada by age")
```

## Adding in migration

Create a Leslie matrix for Canada using the 2010 WPP data. Here's the function:

```{r}
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=17,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  L[1,ncol(L)] <- 0
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```
Here's the matrix:

```{r}
nLx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada", age<85) %>% 
  select(Lx) %>% pull()

## need to fix first age group

nLx <- c(sum(nLx[1:2]), nLx[3:length(nLx)])

nFx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx = ifelse(is.na(Fx), 0, Fx)) %>% 
  select(Fx) %>% pull()

nFx <- nFx[-1]

A <- leslie(nLx, nFx)
A
```

Get the initial population vector for projection:

```{r}
Kt <- d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<85) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  select(pop) %>% 
  pull()
  
ggplot(tibble(age = seq(0, 80, by = 5), K = Kt), aes(age, K)) + 
  geom_bar(stat= "identity") + coord_flip() + ylab("Population") + ggtitle("Female population, Canada 2010")
```

Do a projection with this Leslie matrix for 40 periods:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- A%*%K[,i-1] 
}

```

Now. Let's add in the migrant schedule. 

```{r}
It <- d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age), pop = as.numeric(pop)) %>% 
  left_join(dmig) %>% 
  mutate(prop = mig/pop) %>% 
  filter(age<85) %>% 
  select(prop) %>% 
  pull()
```


```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kmig <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kmig[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kmig[,i] <- A%*%(Kmig[,i-1]*(1+It/2)) + (Kmig[,i-1]*(It/2))
}
```

What difference does adding in migration make?

```{r}
dproj <- tibble(age = age_groups, closed = K[, ncol(K)], with_migration = Kmig[, ncol(K)])

dproj %>%
  gather(projection, pop, -age) %>% 
  group_by(projection) %>% 
  mutate(prop = pop/sum(pop)) %>% 
  ggplot(aes(age, prop, fill = projection)) + geom_bar(stat = 'identity', position = 'dodge') + 
  coord_flip() + 
  ylab("proportion") + 
  ggtitle("Canada's population age structure, 2100\nwith and without migration")
  ggsave(here("plots", "CAN_proj_close_mig.pdf"))
```

## Effect of fertility and mortality on population age strcture

Calculate the TFR for Canada. It's 1.61, which is low. What would happen if we increased the TFR to 2? Let's investigate compared to increasing migration. 

```{r}
TFR <- df %>% 
  filter(region=="Canada", year==2010) %>% 
  summarise(TFR = sum(Fx)*5) %>% 
  pull()

TFR
```

Create new fertility rates:

```{r}

dfert <- df %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(Fx_high = Fx*2.1/TFR)

```

Create a new Leslie Matrix based on these rates:

```{r}
nFx <- dl %>% 
  left_join(dfert) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx_high = ifelse(is.na(Fx_high), 0, Fx_high)) %>% 
  select(Fx_high) %>% pull()

nFx <- nFx[-1]

A1 <- leslie(nLx, nFx)
A1
```

This corresponds to increasing fertility by 25%. Compare the projection results if migration was increased by 25%. 

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kfh <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kfh[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kfh[,i] <- A1%*%(Kfh[,i-1]*(1+It/2)) + (Kfh[,i-1]*(It/2))
}
```

```{r}
Ith <- It*c(rep(0, 3), rep(2.1/TFR, 7), rep(0, 7))

age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kmh <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kmh[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kmh[,i] <- A%*%(Kmh[,i-1]*(1+Ith/2)) + (Kmh[,i-1]*(Ith/2))
}
```

Plot the differences: 
```{r}
dproj <- tibble(age = age_groups, base = Kmig[, ncol(K)], 
                high_fertility = Kfh[, ncol(K)],
                high_migration = Kmh[, ncol(K)])

dproj %>%
  gather(projection, pop, -age) %>% 
  group_by(projection) %>% 
  mutate(prop = pop/sum(pop)) %>% 
  ggplot(aes(age, prop, fill = projection)) + geom_bar(stat = 'identity', position = 'dodge') + 
  facet_grid(~projection) + 
  coord_flip() + 
  ylab("proportion") + 
  ggtitle("Canada's population age structure, 2100\nBaseline, and fertility and migration increased by 25%")
  ggsave(here("plots", "CAN_proj_scenarios.pdf"))
```


# Log-linear models

Load in US regional data from ACS 2017:

```{r}
dr <- readRDS(here("data", "US_region_mig.RDS"))
head(dr)

dr %>% 
  spread(to_region, n)

dr <- dr %>% 
  mutate(n_nd = ifelse(from_region==to_region, 0, n))

dr %>% 
  ungroup() %>% 
  summarise(sum(n_nd))
```

```{r}

mod <- glm(n_nd~as.factor(from_region)+as.factor(to_region),poisson(link=log), data = dr %>% filter(n_nd>0))

summary(mod)

dr %>% 
  ungroup() %>% 
  filter(n_nd>0) %>% 
  select(-n) %>%
  rename(data = n_nd) %>% 
  bind_cols(fitted = fitted(mod)) %>% 
  mutate(from_region =
               factor(
                   from_region,
                   levels = 1:4,
                   labels = c("Northeast", "Midwest", "South", "West")
               ),
         to_region =
               factor(
                   to_region,
                   levels = 1:4,
                   labels = c("Northeast", "Midwest", "South", "West")
                   )) %>% 
  gather(type, migrants, -to_region, -from_region) %>% 
  ggplot(aes(from_region, to_region, fill = migrants/1000)) +
  geom_tile() +
  facet_grid(~type) +
  theme_bw() + 
  scale_fill_viridis_c() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  ggtitle("Inter-regional migration, US 2016-2017\n Data from ACS, fitted with log-linear model")
ggsave(here("plots", "US_regions_mig.pdf"))
```

# Model age curves of migration

```{r}
mig <- read_rds(here("data","US_age_mig.RDS"))

y <- mig %>% ungroup() %>% filter(migrate1==3, statefip==12) %>%  select(prop) %>% pull()
x <- as.numeric(as.character((unique(mig$age_group))))

mig_data <- list(
  N = length(x),
  y=y,
  x = x
)

fit1 <- stan(
  file = here("code", "rc.stan"),  # Stan program
  data = mig_data,    # named list of data
  chains = 4,             # number of Markov chains
  warmup = 1000,          # number of warmup iterations per chain
  iter = 3000,            # total number of iterations per chain
  cores = 2,              # number of cores (using 2 just for the vignette)
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  refresh = 2000          # show progress every 'refresh' iteration
)

fit1


```


```{r}

list_of_draws <- rstan::extract(fit1)

y_hat <- matrix(nrow = length(list_of_draws[[1]]), ncol = length(x))
for(i in 1:length(x)){
  y_hat[,i] <- list_of_draws[["a1"]]*exp(-1*list_of_draws[["alpha1"]]*x[i]) + 
    list_of_draws[["a2"]]*exp(-1*list_of_draws[["alpha2"]]*(x[i] - list_of_draws[["mu2"]]) - exp(-1*list_of_draws[["gamma2"]]*(x[i] - list_of_draws[["mu2"]]))) +
    list_of_draws[["a3"]]*exp(-1*list_of_draws[["alpha3"]]*(x[i] - list_of_draws[["mu3"]]) - exp(-1*list_of_draws[["gamma3"]]*(x[i] - list_of_draws[["mu3"]]))) +
    list_of_draws[["c"]]
}

dfit <- tibble(age = x, 
               data = y, median = apply(y_hat, 2, median),
               lower = apply(y_hat, 2, quantile,0.025),
               upper = apply(y_hat, 2, quantile, 0.975))

dfit %>% 
  ggplot(aes(age, data)) + 
  geom_point(color = "red") + 
  geom_line(aes(x = age, y = median)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
  ggtitle("Florida migration age schedule, 2017 \nData and Rogers-Castro fit")
ggsave(here("plots", "FL_age_mig_rc.pdf"))  
```

