---
title: 'Week 5: Migration'
author: "Monica Alexander"
date: "2/07/2019"
output: html_document
---

Code to produce plots in lecture notes for week 5. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r}
library(tidyverse)
library(here)
library(rstan)
```

# Load in WPP data

Load in fertility rates and tidy up:

```{r}
df <- read_csv(here("data", "WPP2017_FERT_F07_AGE_SPECIFIC_FERTILITY.csv"), skip = 12)

df <- df %>% 
  rename(region = `Region, subregion, country or area *`, period = Period) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  mutate(year = as.numeric(substr(period, 1, 4))) %>% 
  gather(age, Fx, -region, -period, -year) %>% 
  mutate(age = as.numeric(age), Fx = Fx/1000)

head(df)
```


Load in the life table data and tidy up:

```{r}
dl <- read_csv(here("data", "WPP2017_MORT_F17_3_ABRIDGED_LIFE_TABLE_FEMALE.csv"), skip = 16)

dl <- dl %>% 
  rename(region = `Region, subregion, country or area *`,
         Lx = `Number of person-years lived L(x,n)`,
         age = `Age (x)`,
         period = Period) %>% 
  select(region, period, age, Lx) %>% 
  mutate(year = as.numeric(substr(period, 1, 4)), Lx = Lx/10^5)

head(dl)
```

Load in the population data: 

```{r}
d_female <- read_csv(here("data", "WPP2017_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.csv"), skip = 12)

d_female <- d_female %>% 
  rename(region = `Region, subregion, country or area *`,
         year = `Reference date (as of 1 July)`) %>% 
  select(-Index, -Variant, -Notes, -`Country code`) %>% 
  gather(age, pop, -region, -year)

head(d_female)
```

# Population pyramids for Canada

For Canada we have some age-specific data from StatCan:

```{r}

dmig <- read_csv(here("data", "CAN_mig_age.csv"))

dmig <- dmig %>% mutate(mig = mig/1000)
head(dmig)
```

Let's plot as a proportion of total population:

```{r}

d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age), pop = as.numeric(pop)) %>% 
  left_join(dmig) %>% 
  mutate(prop = mig/pop) %>% 
  ggplot(aes(age, prop)) + geom_line() + 
  ggtitle("Migration into Canada by age")
```

## Adding in migration

Create a Leslie matrix for Canada using the 2010 WPP data. Here's the function:

```{r}
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=17,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  L[1,ncol(L)] <- 0
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```
Here's the matrix:

```{r}
nLx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada", age<85) %>% 
  select(Lx) %>% pull()

## need to fix first age group

nLx <- c(sum(nLx[1:2]), nLx[3:length(nLx)])

nFx <- dl %>% 
  left_join(df) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx = ifelse(is.na(Fx), 0, Fx)) %>% 
  select(Fx) %>% pull()

nFx <- nFx[-1]

A <- leslie(nLx, nFx)
A
```

Get the initial population vector for projection:

```{r}
Kt <- d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age)) %>% 
  filter(age<85) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  select(pop) %>% 
  pull()
  
ggplot(tibble(age = seq(0, 80, by = 5), K = Kt), aes(age, K)) + 
  geom_bar(stat= "identity") + coord_flip() + ylab("Population") + ggtitle("Female population, Canada 2010")
```

Do a projection with this Leslie matrix for 40 periods:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- A%*%K[,i-1] 
}

```

Now. Let's add in the migrant schedule. Get the migration rates:

```{r}
It <- d_female %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(age = as.numeric(age), pop = as.numeric(pop)) %>% 
  left_join(dmig) %>% 
  mutate(prop = mig/pop) %>% 
  filter(age<85) %>% 
  select(prop) %>% 
  pull()
```

Do the projection:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kmig <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kmig[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kmig[,i] <- A%*%(Kmig[,i-1]*(1+It/2)) + (Kmig[,i-1]*(It/2))
}
```

What difference does adding in migration make?

```{r}
dproj <- tibble(age = age_groups, closed = K[, ncol(K)], with_migration = Kmig[, ncol(K)])

dproj %>%
  gather(projection, pop, -age) %>% 
  group_by(projection) %>% 
  mutate(prop = pop/sum(pop)) %>% 
  ggplot(aes(age, prop, fill = projection)) + geom_bar(stat = 'identity', position = 'dodge') + 
  coord_flip() + 
  ylab("proportion") + 
  ggtitle("Canada's population age structure, 2210\nwith and without migration")
  ggsave(here("plots", "CAN_proj_close_mig.pdf"))
```

## Effect of fertility and mortality on population age strcture

Calculate the TFR for Canada. It's 1.61, which is low. What would happen if we increased the TFR to 2? Let's investigate compared to increasing migration. 

```{r}
TFR <- df %>% 
  filter(region=="Canada", year==2010) %>% 
  summarise(TFR = sum(Fx)*5) %>% 
  pull()

TFR
```

Create new fertility rates:

```{r}

dfert <- df %>% 
  filter(region=="Canada", year==2010) %>% 
  mutate(Fx_high = Fx*2.1/TFR)

```

Create a new Leslie Matrix based on these rates:

```{r}
nFx <- dl %>% 
  left_join(dfert) %>% 
  filter(year==2010, region == "Canada") %>% 
  mutate(Fx_high = ifelse(is.na(Fx_high), 0, Fx_high)) %>% 
  select(Fx_high) %>% pull()

nFx <- nFx[-1]

A1 <- leslie(nLx, nFx)
A1
```

This corresponds to increasing fertility by 25%. Do the projection:

```{r}
age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kfh <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kfh[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kfh[,i] <- A1%*%(Kfh[,i-1]*(1+It/2)) + (Kfh[,i-1]*(It/2))
}
```


Compare the projection results if migration was increased by 25%. 

```{r}
Ith <- It*c(rep(2.1/TFR, 10), rep(0, 7))

age_groups <- seq(0, 80, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kt
# define population matrix K
Kmh <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
Kmh[,1] <- Kt[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  Kmh[,i] <- A%*%(Kmh[,i-1]*(1+Ith/2)) + (Kmh[,i-1]*(Ith/2))
}
```

Plot the differences: 
```{r}
dproj <- tibble(age = age_groups, base = Kmig[, ncol(K)], 
                high_fertility = Kfh[, ncol(K)],
                high_migration = Kmh[, ncol(K)])

dproj %>%
  gather(projection, pop, -age) %>% 
  group_by(projection) %>% 
  mutate(prop = pop/sum(pop)) %>% 
  ggplot(aes(age, prop, fill = projection)) + geom_bar(stat = 'identity', position = 'dodge') + 
  facet_grid(~projection) + 
  coord_flip() + 
  ylab("proportion") + 
  ggtitle("Canada's population age structure, 2210\nBaseline, and fertility and migration increased by 25%")
  ggsave(here("plots", "CAN_proj_scenarios.pdf"))
```

Implied growth rates:

```{r}
(log(sum(Kmig[,41]) ) - log(sum(Kmig[,40]) ))/5
(log(sum(Kfh[,41]) ) - log(sum(Kfh[,40]) ))/5
(log(sum(Kmh[,41]) ) - log(sum(Kmh[,40]) ))/5
```


# Two population Leslie Matrices

Let's try a two population leslie matrix with Tasmania and Victoria. 

Load in the life tables and make into five-year age groups:

```{r}

# VIC
dv <- read_csv(here("data", "VIC_lt.csv"))

dv$age_group <- cut(dv$Age,
                                 breaks= c(seq(0, 75, by = 5), Inf),
                                 labels = seq(0, 75, by = 5),
                                 right = FALSE)
dv$age_group <- as.numeric(as.character(dv$age_group))

dv <- dv %>% 
  group_by(age_group) %>% 
  summarise(Lx = sum(Lx)/10^5) %>% 
  mutate(state="Victoria")

# TAS

dt <- read_csv(here("data", "TAS_lt.csv"))

dt$age_group <- cut(dt$Age,
                                 breaks= c(seq(0, 75, by = 5), Inf),
                                 labels = seq(0, 75, by = 5),
                                 right = FALSE)
dt$age_group <- as.numeric(as.character(dt$age_group))

dt <- dt %>% 
  group_by(age_group) %>% 
  summarise(Lx = sum(Lx)/10^5) %>% 
  mutate(state="Tasmania")

dlt <- bind_rows(dv, dt)
head(dlt)
```

Load in the fertility data:

```{r}
dfa <- read_csv(here("data", "AUS_state_fertility.csv"))
head(dfa)

dfa <- dfa %>% 
  filter(Measure=="Fertility rate", 
         Region %in% c("Victoria", "Tasmania")) %>% 
  select(Age, Region, Value) %>% 
  mutate(age_group=as.numeric(substr(Age, 1,2))) %>% 
  rename(region=Region, Fx = Value) %>% 
  select(-Age, age_group, region, Fx)
```

Load in population data:

```{r}
dap <- read_csv(here("data", "AUS_state_pop.csv"))

# plot age pyramids

dap %>% 
  ungroup() %>% 
  filter(State %in% c("Australia", "Tasmania"), Measure=="Estimated Resident Population") %>% 
  select(State, Age, Sex, Value) %>% 
  rename(region = State, pop = Value) %>% 
  mutate(age_group = as.numeric(substr(Age, 1,2))) %>% 
  filter(age_group<75, Sex!="Persons") %>% 
  arrange(age_group) %>% 
  group_by(region) %>% 
  mutate(prop = pop/sum(pop),
         prop=ifelse(Sex=="Males", -prop, prop)) %>% 
  ggplot(aes(x = age_group, y = prop, fill = Sex)) + 
  facet_wrap(~region)+
  geom_bar(stat="identity")+
  ggtitle("Population in each age group")+
  ylab("proportion")+ 
  coord_flip() + 
  scale_y_continuous(breaks = seq(-.1, .1, 0.025), 
                     labels = c(seq(.1, 0, -0.025), seq(0.025, .1, 0.025))) + 
  scale_fill_brewer(palette = "Set1") 
ggsave(here("plots", 
            "AUS_TAS_pyramid.pdf"))


dap <- dap %>% 
  filter(State %in% c("Victoria", "Tasmania"), Sex == "Females", Measure=="Estimated Resident Population") %>% 
  select(State, Age, Value) %>% 
  rename(region = State, pop = Value) %>% 
  mutate(age_group = as.numeric(substr(Age, 1,2))) 
head(dap)
```

Load in the migration data:

```{r}
dma <- read_csv(here("data", "AUS_interstate_mig.csv"))
head(dma)

dma <- dma %>% 
  filter(Region %in% c("Victoria", "Tasmania"), Sex == "Females") %>% 
  select(Age, Region, Value, `Interstate Migration`) %>% 
  mutate(age_group=as.numeric(substr(Age, 1,2))) %>% 
  rename(region=Region)
```

We need to work out about what percent of in and out migration total is due to movements between these two states.

```{r}
dmf <- read_csv(here("data", "AUS_interstate_flows.csv"))

in_mig <- dmf %>% 
  filter(`State of Arrival` %in% c("Victoria", "Tasmania"), Sex == "Females") %>% 
  rename(region_in = `State of Arrival`, region_out = `State of Departure`) %>% 
  group_by(region_in, region_out) %>% 
  summarise(in_mig = sum(Value)) %>% 
  group_by(region_in) %>% 
  mutate(prop_in = in_mig/sum(in_mig)) %>% 
  filter(region_out %in% c("Victoria", "Tasmania"))

out_mig <- dmf %>% 
  filter(`State of Departure` %in% c("Victoria", "Tasmania"), Sex == "Females") %>% 
  rename(region_in = `State of Arrival`, region_out = `State of Departure`) %>% 
  group_by(region_in, region_out) %>% 
  summarise(out_mig = sum(Value)) %>% 
  group_by(region_out) %>% 
  mutate(prop_out = out_mig/sum(out_mig)) %>% 
  filter(region_in %in% c("Victoria", "Tasmania"))

both_mig <- in_mig %>% 
  left_join(out_mig)

both_mig

prop_in_vic_to_tas <- both_mig %>% filter(region_in=="Tasmania", region_out=="Victoria") %>% select(prop_in) %>% pull()
prop_in_tas_to_vic <- both_mig %>% filter(region_in=="Victoria", region_out=="Tasmania") %>% select(prop_in) %>% pull()
prop_out_vic_to_tas <- both_mig %>% filter(region_in=="Tasmania", region_out=="Victoria") %>% select(prop_out) %>% pull()
prop_out_tas_to_vic <- both_mig %>% filter(region_in=="Victoria", region_out=="Tasmania") %>% select(prop_out) %>% pull()

```

27% of in-migration to TAS comes from VIC; 5% of in-migration to VIC comes from TAS. 
5% of out migration from VIC is to TAS. 35% of out migration from TAS is to VIC. 

So want to multiply the age schedules by these numbers:

```{r}
dio <- dma %>% 
  filter(`Interstate Migration`!= "NET Interstate") %>% 
  rename(measure = `Interstate Migration`) %>% 
   mutate(mig_adj = case_when(
     region=="Tasmania"&measure=="NIM Arrivals" ~ Value*prop_in_vic_to_tas,
     region=="Tasmania"&measure=="NIM Departures" ~ Value*prop_out_tas_to_vic,
     region=="Victoria"&measure=="NIM Arrivals" ~ Value*prop_in_tas_to_vic,
     region=="Victoria"&measure=="NIM Departures" ~ Value*prop_out_vic_to_tas
  ))

dnet <- dio %>% 
  group_by(age_group, region) %>% 
  summarise(net = mig_adj[measure=="NIM Arrivals"] - mig_adj[measure=="NIM Departures"]) 
```

Work out probability of migration based on populations:

```{r}
mig_prop <- dap %>% 
  select(-Age) %>% 
  left_join(dio) %>% 
  mutate(mig_prop = mig_adj/pop) 

mig_prop %>% 
  filter(measure=="NIM Departures") %>% 
  ggplot(aes(age_group, mig_prop, color = region)) + geom_line() +
  ggtitle("Proportion of population migrating by age")+
  ylab("proportion") + xlab("age")
ggsave(here("plots", "TAS_VIC_mig_age.pdf"))
```

Set up matrix as Vic, Tas to Vic, Tas, Vic to Tas. Assuming return migration, there are ten separate matrices. The native ones are as before with extra multipliers based on probability of migration

```{r}
Lx_vic <- dlt %>% filter(state=="Victoria") %>% filter(age_group<75) %>% select(Lx) %>% pull()
Lx_tas <- dlt %>% filter(state=="Tasmania") %>% filter(age_group<75) %>% select(Lx) %>% pull()

Fx_vic <- c(rep(0, 3), dfa %>% filter(region=="Victoria")  %>% select(Fx) %>% pull(), rep(0, 5))/1000
Fx_tas <- c(rep(0, 3), dfa %>% filter(region=="Tasmania")  %>% select(Fx) %>% pull(), rep(0, 5))/1000

Mx_vic <- mig_prop %>% arrange(age_group) %>% filter(region=="Victoria", measure=="NIM Departures") %>% filter(age_group<75) %>% select(mig_prop) %>% pull()
Mx_tas <- (mig_prop %>% arrange(age_group) %>% filter(region=="Tasmania", measure=="NIM Departures") %>% filter(age_group<75) %>% select(mig_prop) %>% pull())*3

Mx_vic_return <- rep(0.8, length(Mx_tas))
Mx_tas_return <- rep(0.1, length(Mx_tas))

```


```{r}
A_vn_vn <- leslie(nLx = Lx_vic, nFx = Fx_vic, n_age_groups = length(Lx_vic))*(1-Mx_vic)
A_tn_tn <- leslie(nLx = Lx_tas, nFx = Fx_tas, n_age_groups = length(Lx_vic))*(1-Mx_tas)

A_vn_tm <- leslie(nLx = Lx_tas, nFx = Fx_tas, n_age_groups = length(Lx_vic))*(Mx_vic)
A_tn_vm <- leslie(nLx = Lx_vic, nFx = Fx_vic, n_age_groups = length(Lx_vic))*(Mx_tas)

A_tm_tm <- leslie(nLx = Lx_tas, nFx = rep(0, length(Lx_vic)), n_age_groups = length(Lx_vic))*(1-Mx_vic_return)
A_vm_vm <- leslie(nLx = Lx_vic, nFx = rep(0, length(Lx_vic)), n_age_groups = length(Lx_vic))*(1-Mx_tas_return)

A_tm_tn <- leslie(nLx = Lx_tas, nFx = Fx_tas, n_age_groups = length(Lx_vic))
A_tm_tn <- rbind(A_tm_tn[1,], matrix(0, nrow = (length(Lx_vic)-1), ncol = length(Lx_vic)))*(1-Mx_tas_return)
A_vm_vn <- leslie(nLx = Lx_vic, nFx = Fx_vic, n_age_groups = length(Lx_vic))
A_vm_vn <- rbind(A_vm_vn[1,], matrix(0, nrow = (length(Lx_vic)-1), ncol = length(Lx_vic)))*(1-Mx_vic_return)

A_vm_return <- leslie(nLx = Lx_vic, nFx = Fx_vic, n_age_groups = length(Lx_vic))*(Mx_vic_return)
A_tm_return <- leslie(nLx = Lx_tas, nFx = Fx_tas, n_age_groups = length(Lx_vic))*(Mx_tas_return)

```


```{r}
log(Re(eigen(A_vn_vn)$values[1]))/5
log(Re(eigen(A_tn_tn)$values[1]))/5
```

Create one big Leslie matrix:

```{r}
M <- rbind(cbind(A_vn_vn, A_vm_vn, matrix(0, 15, 15), matrix(0, 15, 15)),
cbind(matrix(0, 15, 15), A_vm_vm, A_tn_vm, A_vm_return),
cbind(matrix(0, 15, 15), matrix(0, 15, 15), A_tn_tn, A_tm_tn),
cbind(A_vn_tm, A_tm_return, matrix(0, 15, 15), A_tm_tm))
```

Set up initial population:

```{r}
K_vic <- dap %>% 
  filter(age_group<75, region=="Victoria") %>% 
  arrange(age_group) %>% 
  select(pop) %>% 
  pull()

K_tas <- dap %>% 
  filter(age_group<75, region=="Tasmania") %>% 
  arrange(age_group) %>% 
  select(pop) %>% 
  pull()

M_vic_to_tas <- mig_prop %>% 
  filter(region=="Victoria", measure=="NIM Departures") %>%
  filter(age_group<75) %>% 
  arrange(age_group) %>% 
  select(mig_adj) %>% 
  pull()

M_tas_to_vic <- mig_prop %>% 
  filter(region=="Tasmania", measure=="NIM Departures") %>%
  filter(age_group<75) %>% 
  arrange(age_group) %>% 
  select(mig_adj) %>% 
  pull()
  
Kall <- c(K_vic, M_tas_to_vic, K_tas, M_vic_to_tas)
Kall
```

Project

```{r}
age_groups <- seq(0, 70, by = 5)
n_age_groups <-  length(age_groups)
n_projections <- 40
initial_pop <- Kall
# define population matrix K
K2 <- matrix(0, nrow = n_age_groups*4, ncol = n_projections+1)
K2[,1] <- Kall

# do the projection!
for(i in 2:(n_projections+1)){
  K2[,i] <- M%*%(K2[,i-1])
}
```

```{r}
proj_vn <- K2[1:15,]
proj_vm <- K2[16:30,]
proj_tn <- K2[31:45,]
proj_tm <- K2[46:60,]
```
Plot the results:

```{r}

df_proj_tn <- as_tibble(proj_tn)
colnames(df_proj_tn) <- seq(2016, (2016+40*5), by = 5)
df_proj_tn %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  ggplot(aes(as.numeric(year),  pop, color = factor(age))) + geom_line()

df_proj_tm <- as_tibble(proj_tm)
colnames(df_proj_tm) <- seq(2016, (2016+40*5), by = 5)
df_proj_tm %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  group_by(year) %>% 
  mutate(prop = pop/sum(pop), age = factor(age)) %>% 
  ungroup() %>% 
  mutate(year = as.numeric(year)) %>% 
  ggplot(aes(year,  prop, color = age)) + geom_line(lwd = 0.9) + 
  ggtitle("Projected population by age, Tasmanian migrants")
ggsave(here("plots", "TAS_mig_proj.pdf"))

df_proj_vn <- as_tibble(proj_vn)
colnames(df_proj_vn) <- seq(2016, (2016+40*5), by = 5)
df_proj_vn %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  ggplot(aes(as.numeric(year),  pop, color = factor(age))) + geom_line()

df_proj_vm <- as_tibble(proj_vm)
colnames(df_proj_vm) <- seq(2016, (2016+40*5), by = 5)
df_proj_vm %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  group_by(year) %>% 
  mutate(prop = pop/sum(pop), age = factor(age)) %>% 
  ungroup() %>% 
  mutate(year = as.numeric(year)) %>% 
  ggplot(aes(year,  prop, color = age)) + geom_line(lwd = 0.9) + 
  ggtitle("Projected population by age, Victorian migrants")
ggsave(here("plots", "VIC_mig_proj.pdf"))

```

Stable age structures

```{r}
dfproj <- df_proj_tn %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  mutate(type = "Tasmanian native") %>% 
  bind_rows(df_proj_tm %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  mutate(type = "Tasmanian migrant")) %>% 
  bind_rows(df_proj_vn %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  mutate(type = "Victorian native")) %>% 
  bind_rows(df_proj_vm %>% 
  mutate(age = age_groups) %>% 
  gather(year, pop, -age) %>% 
  mutate(type = "Victorian migrant"))

dfproj %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year==2016+n_projections*5) %>% 
  group_by(type) %>% 
  mutate(prop = pop/sum(pop)) %>% 
  ggplot(aes(age, prop, fill = type)) + geom_bar(stat = 'identity', position = 'dodge') + coord_flip() + facet_wrap(~type)
ggsave(here("plots", "TAS_VIC_stable.pdf"))

```


# Log-linear models

Load in US regional data from ACS 2017:

```{r}
dr <- readRDS(here("data", "US_region_mig.RDS"))
head(dr)

dr %>% 
  spread(to_region, n)

dr <- dr %>% 
  mutate(n_nd = ifelse(from_region==to_region, 0, n))

dr %>% 
  ungroup() %>% 
  summarise(sum(n_nd))
```
Fit a log-linear model:
```{r}

mod <- glm(n_nd~as.factor(from_region)+as.factor(to_region),poisson(link=log), data = dr %>% filter(n_nd>0))

summary(mod)

dr %>% 
  ungroup() %>% 
  filter(n_nd>0) %>% 
  select(-n) %>%
  rename(data = n_nd) %>% 
  bind_cols(fitted = fitted(mod)) %>% 
  mutate(from_region =
               factor(
                   from_region,
                   levels = 1:4,
                   labels = c("Northeast", "Midwest", "South", "West")
               ),
         to_region =
               factor(
                   to_region,
                   levels = 1:4,
                   labels = c("Northeast", "Midwest", "South", "West")
                   )) %>% 
  gather(type, migrants, -to_region, -from_region) %>% 
  ggplot(aes(from_region, to_region, fill = migrants/1000)) +
  geom_tile() +
  facet_grid(~type) +
  theme_bw() + 
  scale_fill_viridis_c() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  ggtitle("Inter-regional migration, US 2016-2017\n Data from ACS, fitted with log-linear model")
ggsave(here("plots", "US_regions_mig.pdf"))
```

# Model age curves of migration

This is based on Florida migration (taken from the latest ACS).

```{r}
mig <- read_rds(here("data","US_age_mig.RDS"))

y <- mig %>% ungroup() %>% filter(migrate1==3, statefip==12) %>%  select(prop) %>% pull()
x <- as.numeric(as.character((unique(mig$age_group))))

mig_data <- list(
  N = length(x),
  y=y,
  x = x
)

fit1 <- stan(
  file = here("code", "rc.stan"),  # Stan program
  data = mig_data,    # named list of data
  chains = 4,             # number of Markov chains
  warmup = 1000,          # number of warmup iterations per chain
  iter = 3000,            # total number of iterations per chain
  cores = 2,              # number of cores (using 2 just for the vignette)
  #control = list(adapt_delta = 0.99, max_treedepth = 15),
  refresh = 2000          # show progress every 'refresh' iteration
)

fit1


```
Plot the data, fit and 95% CIs. 

```{r}

list_of_draws <- rstan::extract(fit1)

y_hat <- matrix(nrow = length(list_of_draws[[1]]), ncol = length(x))
for(i in 1:length(x)){
  y_hat[,i] <- list_of_draws[["a1"]]*exp(-1*list_of_draws[["alpha1"]]*x[i]) + 
    list_of_draws[["a2"]]*exp(-1*list_of_draws[["alpha2"]]*(x[i] - list_of_draws[["mu2"]]) - exp(-1*list_of_draws[["gamma2"]]*(x[i] - list_of_draws[["mu2"]]))) +
    list_of_draws[["a3"]]*exp(-1*list_of_draws[["alpha3"]]*(x[i] - list_of_draws[["mu3"]]) - exp(-1*list_of_draws[["gamma3"]]*(x[i] - list_of_draws[["mu3"]]))) +
    list_of_draws[["c"]]
}

dfit <- tibble(age = x, 
               data = y, median = apply(y_hat, 2, median),
               lower = apply(y_hat, 2, quantile,0.025),
               upper = apply(y_hat, 2, quantile, 0.975))

dfit %>% 
  ggplot(aes(age, data)) + 
  geom_point(color = "red") + 
  geom_line(aes(x = age, y = median)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
  ggtitle("Florida migration age schedule, 2017 \nData and Rogers-Castro fit")
ggsave(here("plots", "FL_age_mig_rc.pdf"))  
```

