---
title: "Bayesian Lee-Carter"
author: "Monica Alexander"
date: "7/23/2020"
output: pdf_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's estimate a Lee-Carter model for Australian males. 

NOTE: Bare bones here, will fill in later

# Data

Read in the data and tidy up

```{r}
library(tidyverse)
library(here)
library(janitor) 
library(patchwork)
library(rstan)
library(tidybayes)

d <- read_table(here("data/AUS_Mx_1x1.txt"), skip = 2, col_types = 'dcddd')

# tidy up 
d <- d %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<101) %>% # older ages problematic in earlier years
  pivot_longer(female:total, names_to = "sex", values_to = "mx")

```

# SVD

Get the matrix of age-specific rates:

```{r}
m_tx <- d %>% 
  filter(sex=="male") %>% 
  select(-sex) %>% 
  pivot_wider(names_from = "age", values_from = "mx") %>% 
  select(-year) %>% 
  as.matrix()

ages <- unique(d$age)
years <- unique(d$year)

```

log and demean those rates, and get $a_x$, which is the logged mean:

```{r}
logm_tx <- log(m_tx)
ax <- apply(logm_tx, 2, mean)
```

Do the SVD

```{r}
swept_logm_tx <- sweep(logm_tx, 2, ax)

svd_mx <- svd(swept_logm_tx)

bx <- svd_mx$v[, 1]/sum(svd_mx$v[, 1])
kt <- svd_mx$d[1] * svd_mx$u[, 1] * sum(svd_mx$v[, 1])

```

Plot the results:

```{r}
lc_age_df <- tibble(age = ages, ax = ax, bx = bx)
lc_time_df <- tibble(year = years, kt = kt)

p1 <- ggplot(lc_age_df, aes(age, ax)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("ax values")

p2 <- ggplot(lc_age_df, aes(age, bx)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("bx values")

p3 <- ggplot(lc_time_df, aes(year, kt)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("kt values")

p1+p2+p3
```

What is the interpretation of each of these components? 

<p style="color:red"> To fill in, for now can talk about it </p>

Let's plot the data versus fit for a few different years

```{r, fig.width=8, fig.height=4}
d %>% 
  filter(sex=="male") %>% 
  left_join(lc_age_df) %>% 
  left_join(lc_time_df) %>% 
  mutate(estimate = exp(ax+bx*kt)) %>% 
  select(year, age, mx, estimate) %>% 
  rename(data = mx) %>% 
  #pivot_longer(data:estimate, names_to = "source", values_to = "value") %>% 
  filter(year %in% seq(1928, 2018, by = 30)) %>% 
  ggplot(aes(age, data)) + 
  geom_point(aes(color = "data")) + 
  geom_line(aes(age, estimate, color = "fit"), lwd = 1.1)+
  facet_grid(~year) + 
  scale_y_log10() + 
  ylab("mortality rate") +
  scale_color_brewer(name = "",palette = "Set1") + 
  theme_bw()
```

# Bayesian time series model for $k_t$

Read in deaths and populatioon files

```{r}
d_deaths <- read_table(here("data/AUS_Deaths_1x1.txt"), skip = 2)
d_pop <- read_table(here("data/AUS_Population.txt"), skip = 2)

# tidy up 
d_deaths <- d_deaths %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<101) %>% # older ages problematic in earlier years
  pivot_longer(female:total, names_to = "sex", values_to = "deaths")

d_pop <- d_pop %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<101) %>% # older ages problematic in earlier years
  pivot_longer(female:total, names_to = "sex", values_to = "population")


```

Stan format

```{r}
years_lc <- 2010:2018

D <- d_deaths %>% 
  filter(sex=="male", year %in% years_lc) %>% 
  select(-sex) %>% 
  mutate(deaths = round(deaths)) %>% 
  pivot_wider(names_from = "age", values_from = "deaths") %>% 
  select(-year) %>% 
  as.matrix()

P <- d_pop %>%
  filter(sex=="male", year %in% years_lc) %>%
  select(-sex) %>%
  mutate(population = round(population)) %>%
  pivot_wider(names_from = "age", values_from = "population") %>%
  select(-year) %>%
  as.matrix()


stan_data <- list(D = D, P = P, 
                  nyears = length(years_lc), 
                  nages = length(ages),
                  nprojyears = 5,
                  ax = ax,
                  bx = bx,
                  kt = kt[which(years %in% years_lc)])

mod <- stan(data = stan_data, file = here("code/lee_carter.stan"))
```
Plot $k_t$ projection
```{r}
mod %>% 
  gather_draws(mu_k[i], mu_k_proj[j]) %>% 
  median_qi() %>% 
  ungroup %>% 
  mutate(year = years_lc[1]:2023, kt = c(kt[which(years %in% years_lc)], rep(NA, 5))) %>% 
  ggplot(aes(year, .value)) + geom_line() + 
  geom_point(aes(year, kt))+
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.2)
```

Plot 2018 and 2013 mortality rate estimate/projections

```{r}
d_res <- mod %>% 
  gather_draws(log_mu[t,x], log_mu_proj[p,x]) %>% 
  median_qi() %>% 
  ungroup() %>% 
  mutate(year = years_lc[t]) %>% 
  mutate(year = ifelse(is.na(year), (2019:2023)[p], year)) %>% 
  mutate(age = ages[x]) %>% 
  left_join(d %>% filter(sex=="male") %>% select(year, age, mx)) 

d_res %>% 
  filter(year==2018|year==2023) %>% 
  ggplot(aes(age, exp(.value))) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(.lower), ymax = exp(.upper)), alpha = 0.2) + 
  scale_y_log10() + 
  facet_wrap(~year) + 
  ylab("mortality rate")
```


To-do: convert to life expectancy


